Efficient and Scalable Bipartite Matching with Fast Beta Linkage (fabl)

### 2021-10-11 methods_paper/fabl-paper.tex
edited Efficiency section

To increase the speed of our Gibbs sampler and expand the scale of the linkage tasks we can undertake, we employ hashing methods similar to those used by \cite{enamorado2019using} in the creation of \texttt{fastlink}, a fast and scalable implementation of the Fellegi Sunter method. The key insight is to recognize that record pairs contribute to posterior calculations only through the agreement pattern of the $\gamma_{ij}$ vector. To make this more precise, let 
\(\mathcal{H}\) be the set of unique agreement patterns in the data, let
\(P = |\mathcal{H}|\) denote the total number of unique agreement
patterns.  Observe that \(P\) is bounded above by \(\prod_{f=1}^F L_f\), and that this bound does not depend on \(n_1\) or \(n_2\).  Prior to
processing the data, we identify all \(P\) patterns in \(\mathcal{H}\)
and enumerate them \(1, \ldots, P\), which are called \emph{hashed values}. Additionally, we store ``one-hot encodings'' of these patterns rather than the original Fellegi Sunter comparison vectors, to aid in vectorized computations. Whenever possible, we conduct calculations over these \(P\) agreement
patterns, instead of the typical \(n_1 \times n_2\) record pairs.

@@

This function maps each agreement pattern to a unique integer, allowing us to store a scalar quantity instead of an entire vector for each record pair. For computational ease, we then map these integers to sequential integers from $1, \ldots, P$ corresponding to the enumerated patterns $\mathcal{H}$. When the $(i, j)$ pair exhbits the $p^{th}$ pattern, we say $h(i,j) = p$.

With all record pairs converted to hashed values, we can create a more compact nested list 
$$\mathcal{R} = \left\{\{r_{j_p} \}_{p=1}^{P} \right\}_{j = 1}^{n_2}$$
where $r_{j_p} = \{i \in \bm{X_1} | h(i,j) = p\}$. Note that the pair $(\mathcal{H}, \mathcal{R})$ fully characterizes the comparison matrix $\Gamma$ with no loss of information. This representation is useful because it allows us to easily compute sufficient statistics
$$H_{j_p} = \sum_{i} I(h(i,j) = p) = ||r_{j_p}||$$
and
$$H_p =  \sum_{i, j} I(h(i,j) = \sum_{j} H_{j_p}$$
As will be shown in Section \ref{efficient-posterior}, all posterior calculations are conducted through these sufficient statistics.

@@

\underline{Sample $\Phi^{(s+1)}|\mathcal{H}, \mathcal{R}, \mathbf{Z}^{(s)}$:} Calculated at the level of the record pairs, updating  \(\alpha_{fl}(\mathbf{Z})\) and \(\beta_{fl}(\mathbf{Z})\) for each field and level in the linkage task constitutes \(2 \times \sum L_f\) many summations over \(n_A \times n_B\) quantities. These are simple calculations, but become computationally burdensome when working on large linkage tasks.

Instead, we use one-hot encodings of the agreement patterns $\mathcal{H}$ for more efficient calculations. Denote
$H_p^m = \sum_{j=1}^{n_B} \mathbf{1}_{(Z_j, j) \in h_p}$ to be the
number of matching record pairs with agreement pattern $p$. We can also easily calculate the number of nonmatching record pairs with agreement pattern $p$ as \(H_p^u = H_p - H_p^m\). Then, if $\mathbf{\alpha_0}$ and $\mathbf{\beta_0}$ are vectors of prior parameters for the $\mathbf{m}$ and $\mathbf{u}$ distributions respectively, the posterior update becomes simply

$$\alpha(\mathbf{Z}) = \mathbf{\alpha_0} + \sum_{p=1}^P H_p^m \times h_p$$
$$\beta(\mathbf{Z}) = \mathbf{\beta_0} + \sum_{p=1}^P H_p^u \times h_p$$

@@

\underline{Sample $\mathbf{Z}^{(s+1)}|\mathcal{H}, \mathcal{R}, \Phi^{(s+1)}$:} Although sampling \(Z_j\) from a the
full conditional provided earlier is conceptually straightforward, it
becomes computational burdensome when \(n_A\) is larger. The reader can
confirm that sampling a value from a large set of unequal probabilities
becomes difficult in most programming languages. To speed up
computation, we break this sampling step into two simpler steps. First,
we calculate the Fellegi Sunter weight \(w_{p}\) associated with each
unique pattern and sample the agreement pattern between \(j\) and its
potential match. Second, we sample the record label uniformly among
records associated with that agreement pattern for that particular
\(j\in B\). More concretely, define \(h(Z_j)\) to be the agreement
pattern between \(j\) and its potential match, and say
\(h(Z_j) = h_{P+1}\) when \(Z_j = n_A + 1\). Then,

\[P\left(h\left(Z_j^{(s+1)}\right) = p | \mathbf{m}, \mathbf{u}, \mathbf{Z^{(s)}}\right) \propto
\begin{cases} 
	w_{p}\times H_{p_j}  & p \leq P; \\
	n_A \frac{n_B - D + \beta_{\lambda}}{D + \alpha_{\lambda}} &   p = P + 1 \\
\end{cases}\]
\[P\left(Z_j^{(s+1)} = i\bigg|\;\; h\left(Z_j^{(s+1)}\right) = p\right) = \begin{cases} 
	\frac{1}{H_{p_j}} & (i, j) \in h_p \\
	0 & \text{otherwise} \\
\end{cases}\]

Lastly, we recognize that all posterior updates are governed by the
agreement patterns of the record pairs rather than the record labels
themselves. Thus we complete the entire Gibbs procedure first at the
level of the \(P\) agreement patterns with the first equation above.
After, we can back-fill the records corresponding to the agreement
patterns through the second equation. Sampling uniformly is
computationally simple even for large sets of candidate records, but this
step can also be parallelized when working with large data.

@@

{
\color{blue}
The computational complexity of \texttt{fabl} is determined by its two steps: constructing comparison vectors and Gibbs sampling. The computational complexity of all pairwise comparisons across files $\mathbf{X}_1$ and $\mathbf{X}_2$ is $O(F n_1 n_2)$. \textcolor{blue}{TALK TO BEKA. I CAN DO THE SAME TRICKS AS TED HERE TO REDUCE COMPLEXITY}. Since \texttt{fabl} is able to compute these comparisons in parallel, we can split these computations across $B$ equally sized partitions, so the complexity becomes $O(\frac{F}{B} n_1 n_2)$.

@@

In this section, we revisit an application to the El Salvadoran Civil War, where we estimate the number of documented identifiable deaths (DID). Though the data files used here are small, this study shows how the computational complexity of \texttt{fabl} depends on the number of unique agreement patterns, and how significant computational gains can be achieved by simplifying the construction of the comparison vectors. This case study also considers the ramifications of 
independently sampling $Z_j$ rather than strictly enforcing one-to-one matching as done by \texttt{BRL}.

%The second case study is the National Long Term Care Survey (NLTCS), a much larger linkage tasks which \texttt{BRL} is unable to complete but which we complete with \texttt{fabl} with relative ease.

@@

{
\color{blue} For Beka: You wrote that you wanted more detail here, and I added some of it (with some citations). I believe I have all the important informatoin, just more concise. If I included Sadinle's level of detail, I worried about plagarism.
\vspace{1em}
}

@@

{
\color{blue}
Through \texttt{fabl}, we arrive at a Bayes estimate of 179 individuals recorded in both datasets. We calcuate posterior samples of size of the overlap across files by summing the number of matches found in each iteration of the Gibbs sampler, and subtracting the number of matches that violate one-to-one matching. The posterior 95\% credible interval for the overlap across files is (206, 238), indicating that the Bayes estimate is quite conservative. This is because there a large number of records in ERTL for which there are multiple plausible matches in CDHES; \texttt{fabl} recognizes that a match exists among the several options, but is unable to definitely declare a specific pair a match in the Bayes estimate. We also compute a partial estimate of the linkage structure, using $\theta_{10} = \theta_{01} = 1$, $\theta_{11} = 2$, and $\theta_R = 0.1$ as in the simulation study. Here, the Bayes estimate provides 136 matches of which the model is quite confident, and 175 records to verify manually; this means that after clerical review, the number of individuals replicated across datasets would fall in the interval (136, 311), encapsulating the posterior credible interval. More or fewer records could be indentified for clerical review by decreasing or increasing $\theta_R$.

We see similar results under \texttt{BRL}, with a Bayes estimate of 181 individuals recorded in both datasets, a posterior 95\% credible interval of (211, 244), and a range of (140, 294) after the partial estimate and clerical review. 
}

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/overlap_distribution_smallP_bayes} 
		\caption{Posterior distribution of overlap across the two files. The solid lines show the Bayes estimate for the amount of overlap, and the dashed line is the Bayes estimate under fabl before resolving violations of one-to-one matching}
		\label{fig:overlap-plot}
	\end{center}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/m_posterior_smallP} 
\caption{Posterior estimates of m parameters with 95\% credible intervals}\label{fig:m-and-u}
\label{fig:m-and-u}
\end{center}
\end{figure}

@@

%In this section, we investigate violations of one-to-one matching of our method as compared to \texttt{BRL}.
{
\color{blue}
As noted previously, the application exhibits many situations in which bipartite matching is difficult. To explore further, we calculate the number of matchings throughout the \texttt{fabl} Gibbs sampler that violate our one-to-one assumption, and find on average 26.17 matchings in violation. These matchings are impossible under the \texttt{BRL} Gibbs sampler, so examining these more closely can provide insight how each model handles situations in which one file in \(\bm{X}_1\) has
multiple plausible matches in \(\bm{X}_2\). 
}

@@

%\begin{figure}[t]
%	
%	{\centering \includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/overlap_distribution_smallP_bayes} 
%		
%	}
%	
%	\caption{Posterior distribution of overlap across the two files. The solid lines show the Bayes estimate for the amount of overlap, and the dashed line is the Bayes estimate under fabl before resolving violations of one-to-one matching}\label{fig:overlap-plot}
%\end{figure}

%Overall, both methods present similar results. \texttt{fabl} yields
%an initial Bayes estimate of total 195 DID. After
%resolving matches that violated one-to-one requirements,  we estimate 180 DID. 
%
%
%
%This is very close to the 181 DID found
%by \texttt{BRL}. The main reason for this discrepancy is the difference
%in how each model handles situations in which one file in \(\bm{X}_1\) has
%multiple plausible matches in \(\bm{X}_2\).

%
%\begin{figure}[t]
%	
%	{\centering \includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/rosa_records} 
%		
%	}
%	
%	\caption{Example of linkage situation with multiple plausible matches}\label{fig:rosa-maria}
%\end{figure}

Figure~\ref{fig:rosa-maria} illustrates an example. Note that these
records present a near violation of our assumption that there are no
duplications within files; we continue to assume that records 825 and
826 in ERTL correspond to different individuals, but their records are
nearly identical. In addition, using the modified Levenstein distance, the comparison vectors \(\gamma_{2776, 825}\) and
\(\gamma_{2776, 826}\) are exactly identical.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/rosa_records} 
\caption{Example of linkage situation with multiple plausible matches}\label{fig:rosa-maria}
\end{center}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/bad_mixing} 
\caption{Gibbs sampling in situation with multiple plausible matches.}\label{fig:mixing-plot}
\end{center}
\end{figure}

@@

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/overlap_trace} 
\caption{Traceplot for number of matches found across datasets in El Salvador case study}\label{fig:overlap_trace}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/m_trace} 
\caption{Traceplot for m parameter in El Salvador case study}\label{fig:m_trace}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/u_trace} 
\caption{Traceplot for u parameter in El Salvador case study}\label{fig:u_trace}
\end{center}
\end{figure}

\newpage
\bigskip

>>>>

To increase the speed of our Gibbs sampler and expand the scale of the linkage tasks we can undertake, we employ hashing methods similar to those used by \cite{enamorado2019using} in the creation of \texttt{fastlink}, a fast and scalable implementation of the Fellegi Sunter method. The key insight is to recognize that record pairs contribute to posterior calculations only through the agreement pattern of the $\gamma_{ij}$ vector. To make this more precise, let $h_1, \ldots, h_P$ denote the unique agreement patterns, collect these unique patterns in the set $\mathcal{P} = \{h_1, \ldots, h_P\}$. Here, 
$P = |\mathcal{P}|$ is the total number of unique agreement
patterns; observe that \(P\) is bounded above by \(\prod_{f=1}^F L_f\), and that this bound does not depend on \(n_1\) or \(n_2\). In this context, the integers $1, \ldots, P$ serve as \emph{hashed values} that encode the same information as the orignial vectors themselves. Whenever possible, we conduct calculations over these \(P\) agreement
patterns, instead of the typical \(n_1 \times n_2\) record pairs.

Additionally, we store ``one-hot encodings'' of these patterns rather than the original Fellegi Sunter comparison vectors, to aid in vectorized computations.

@@

This function maps each agreement pattern to a unique integer, allowing us to store a scalar quantity instead of an entire vector for each record pair. For computational ease, we then map these integers to sequential integers from $1, \ldots, P$ corresponding to the enumerated patterns in $\mathcal{P}$. When the $(i, j)$ pair exhibits the $p^{th}$ pattern, we say $(i, j) \in h_p$.

With all record pairs converted to hashed values, we can create a more compact nested list 
$$\mathcal{R} = \left\{\{r_{j_p} \}_{p=1}^{P} \right\}_{j = 1}^{n_2}$$
where $r_{j_p} = \{i \in \bm{X_1} | (i, j) \in h_p\}$. This representation is useful because it allows us to easily compute sufficient statistics
$$H_{j_p} = \sum_{i} I((i, j) \in h_p) = ||r_{j_p}||$$
and
$$H_p =  \sum_{i, j} I((i, j) \in h_p) = \sum_{j} H_{j_p}$$
As will be shown in Section \ref{efficient-posterior}, all posterior calculations are conducted through these sufficient statistics. Note that $\mathcal{P}$ and $\mathcal{R}$ fully characterize the comparison matrix $\Gamma$ with no loss of information, so we proceed with $\tilde{\Gamma} = \{\mathcal{P}, \mathcal{R}\}$ for our efficient posterior inference.

@@

\underline{Sample $\Phi^{(s+1)}|\tilde{\Gamma}, \mathbf{Z}^{(s)}$:} Calculated at the level of the record pairs, updating  \(\alpha_{fl}(\mathbf{Z})\) and \(\beta_{fl}(\mathbf{Z})\) for each field and level in the linkage task constitutes \(2 \times \sum L_f\) many summations over \(n_A \times n_B\) quantities. These are simple calculations, but become computationally burdensome when working on large linkage tasks.

Instead, we use one-hot encodings of the agreement patterns $\mathcal{P}$ for more efficient calculations. Denote
$H_p^m = \sum_{j=1}^{n_B} \mathbf{1}_{(Z_j, j) \in h_p}$ to be the
number of matching record pairs with agreement pattern $p$. We can also easily calculate the number of nonmatching record pairs with agreement pattern $p$ as \(H_p^u = H_p - H_p^m\). Then, if $\mathbf{\alpha_0}$ and $\mathbf{\beta_0}$ are vectors of prior parameters for the $\mathbf{m}$ and $\mathbf{u}$ distributions respectively, the posterior update becomes simply

$$\alpha(\mathbf{Z}) = \bm{\alpha_0} + \sum_{p=1}^P H_p^m \times h_p$$
$$\beta(\mathbf{Z}) = \bm{\beta_0} + \sum_{p=1}^P H_p^u \times h_p$$

@@

\underline{Sample $\mathbf{Z}^{(s+1)}|\tilde{\Gamma}, \Phi^{(s+1)}$:} Although sampling \(Z_j\) from a the
full conditional provided earlier is conceptually straightforward, it
becomes computational burdensome when $n_1$ is large. This is because sampling a value from $n_1$ many options with unequal weights requires normalizing the weights to probabilities, which has a computational cost that scales with $n_1$. To speed up
computation, we break this sampling step into two simpler steps. First,
we calculate the Fellegi Sunter weight \(w_{p}\) associated with each
unique pattern and sample the agreement pattern between \(j\) and its
potential match. Second, we sample the record label uniformly among
records associated with that agreement pattern for that particular
\(j\in B\). More concretely, define \(h(Z_j)\) to be the agreement
pattern between \(j\) and its potential match, and say
\(h(Z_j) = h_{P+1}\) when \(Z_j = n_A + 1\). Then,

$$P\left(h\left(Z_j^{(s+1)}\right) = p \mid \Phi, \mathbf{Z^{(s)}}\right) \propto
\begin{cases} 
	w_{p}\times H_{j_p}  & p \leq P; \\
	n_A \frac{n_B - D + \beta_{\lambda}}{D + \alpha_{\lambda}} &   p = P + 1 \\
\end{cases}$$
$$P\left(Z_j^{(s+1)} = i \mid h\left(Z_j^{(s+1)}\right) = p\right) \begin{cases} 
	\frac{1}{H_{j_p}} & (i, j) \in h_p \\
	0 & \text{otherwise} \\
\end{cases}$$

Lastly, we recognize that all posterior updates are governed by the
agreement patterns of the record pairs rather than the record labels
themselves. Thus we complete the entire Gibbs procedure first at the
level of the \(P\) agreement patterns with the first equation above.
After all Gibbs iterations are complete, we can back-fill the records corresponding to the agreement patterns through the second equation.

We emphasize the computational gains of this split sampler: the first step is a sample from $P$ many options, where $P$ does not scale with the size of the linkage task; and the second step is sampling uniformly at random, which is computationally simple even for large sets of candidate records. These changes in themselves drastically improves the speed of the sampler, and each can be parallelized if desired for additional computational gains.

@@

{
\color{blue}
The computational complexity of \texttt{fabl} is determined by its two steps: constructing comparison vectors and Gibbs sampling. The computational complexity of all pairwise comparisons across files $\mathbf{X}_1$ and $\mathbf{X}_2$ is $O(F n_1 n_2)$. Since \texttt{fabl} is able to compute these comparisons in parallel, we can split these computations across $B$ equally sized partitions, so the complexity becomes $O(\frac{F}{B} n_1 n_2)$.

@@

In this section, we revisit an application to the El Salvadoran Civil War, where we estimate the number of individuals recorded in each dataset. Though the data files used here are small, this study shows how the computational complexity of \texttt{fabl} depends on the number of unique agreement patterns found in the data, and how significant computational gains can be achieved by simplifying the construction of the comparison vectors. This case study also considers the ramifications of independently sampling $Z_j$ rather than strictly enforcing one-to-one matching as done by \texttt{BRL}.

@@

@@

Through \texttt{fabl}, we arrive at a Bayes estimate of 179 individuals recorded in both datasets. We calcuate posterior samples of size of the overlap across files by summing the number of matches found in each iteration of the Gibbs sampler, and subtracting the number of matches that violate one-to-one matching. The posterior 95\% credible interval for the overlap across files is (206, 238), indicating that the Bayes estimate is quite conservative. This is because there a large number of records in ERTL for which there are multiple plausible matches in CDHES; \texttt{fabl} recognizes that a match exists among the several options, but is unable to definitely declare a specific pair a match in the Bayes estimate. We also compute a partial estimate of the linkage structure, using $\theta_{10} = \theta_{01} = 1$, $\theta_{11} = 2$, and $\theta_R = 0.1$ as in the simulation study. Here, the Bayes estimate provides 136 matches of which the model is quite confident, and 175 records to verify manually; this means that after clerical review, the number of individuals replicated across datasets would fall in the interval (136, 311), encapsulating the posterior credible interval. More or fewer records could be indentified for clerical review by decreasing or increasing $\theta_R$.

We see similar results under \texttt{BRL}, with a Bayes estimate of 181 individuals recorded in both datasets, a posterior 95\% credible interval of (211, 244), and a range of (140, 294) after the partial estimate and clerical review.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/overlap_distribution_smallP_bayes} 
		\caption{Posterior distribution and Bayes estimate of overlap across the two files. We note they are quite similar under both methods.}
		\label{fig:overlap-plot}
	\end{center}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/m_posterior_smallP} 
\caption{Posterior estimates of m parameters with 95\% credible intervals. We note they are quite similar across the two methods. }\label{fig:m-and-u}
\label{fig:m-and-u}
\end{center}
\end{figure}

@@

As noted previously, the application exhibits many situations in which bipartite matching is difficult. To explore further, we calculate the number of matchings throughout the \texttt{fabl} Gibbs sampler that violate our one-to-one assumption, and find on average 26.17 matchings in violation. These matchings are impossible under the \texttt{BRL} Gibbs sampler, so examining these more closely can provide insight how each model handles situations in which one file in \(\bm{X}_1\) has
multiple plausible matches in \(\bm{X}_2\).

@@

Figure~\ref{Tab:rosa-maria} illustrates an example. Note that these
records present a near violation of our assumption that there are no
duplications within files; we continue to assume that records 825 and
826 in ERTL correspond to different individuals, but their records are
nearly identical. In addition, using the modified Levenstein distance, the comparison vectors \(\gamma_{2776, 825}\) and
\(\gamma_{2776, 826}\) are exactly identical.

\begin{table}
	\centering
	\begin{tabular}[h!]{llllllll}
		\hline
		Record Number & dataset & lastname & firstname & day & month & year & dept \\
		\hline
		825 & CDHES & Pineda & Rosa & 6 & 4 & 1984 & NA\\
		\hline
		826 & CDHES & Pineda & Rosa Maria & 6 & 4 & 1984 & NA\\
		\hline
		2776 & ERTL & Pineda & Rosa Maria & 4 & 4 & 1984 & NA\\
		\hline
	\end{tabular}
	\caption{Example of linkage situation in which record 2776 in ERTL has multiple plausible matches in CDHES, leading to undesirable behavior both \texttt{fabl} and \texttt{BRL}.} \label{Tab:rosa_maria}
\end{table}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/bad_mixing} 
\caption{Gibbs sampling in situation with multiple plausible matches. Under \texttt{BRL}, two records in ERTL compete for their most similar record, splitting the posterior probability. Under \texttt{fabl}, both record consistently match with their most similar record, consistently violating one-to-one matching.} \label{fig:mixing-plot}
\end{center}
\end{figure}

@@

\clearpage

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/overlap_trace} 
\caption{Traceplot for number of matches found across datasets in El Salvador case study}\label{fig:overlap_trace}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/m_trace} 
\caption{Traceplot for m parameter in El Salvador case study}\label{fig:m_trace}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/u_trace} 
\caption{Traceplot for u parameter in El Salvador case study}\label{fig:u_trace}
\end{center}
\end{figure}

\clearpage

@@
